---
title: "Breakout Activity - Debug Exercise 4"
output: html_document
---

## Breakout Activity - Debug Exercise 4 (Introduction to Spatial Vector Data)

**Groups of 3-4. 5-8 minutes, then brief share-out.**

The code below is the *original* version of the Exercise 4 plotting chunk from the
Introduction to Spatial Vector Data tutorial. It does not run as written. Work through it as a
group, diagnose the error, and fix it.

```{r climate-setup}
# recreates the objects from Exercises 1-4 of the tutorial, so this chunk runs standalone
library(sf)
library(dplyr)

set.seed(809)

# longitude and latitude objects
longitude <- runif(n=10, min=-120, max=-115)
latitude <- runif(n=10, min=35, max=45)

# combine into a matrix, then a data frame
lonlat <- cbind(longitude, latitude)
lonlat.df <- as.data.frame(lonlat)

# convert to a spatial object
lonlat.sf <- sf::st_as_sf(lonlat.df, coords = c("longitude", "latitude"))
sf::st_crs(lonlat.sf) <- 4326

# create the climate data frame and join it to lonlat.sf
climate <- data.frame(temperature = rnorm(n=10, mean=25, sd=5), precip = runif(10, min=0, max=500))
climatePts <- lonlat.sf %>%
  dplyr::mutate(climate)
```

```{r broken-plot, error=TRUE}
# ggplot2
library(ggplot2)

# plot the result
ggplot(climatePts) +
  geom_point(aes(x = longitude, y = latitude, color = temperature, size = precip)) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal()
```

### Discussion questions

1. What error do you get when you run the `broken-plot` chunk?
2. Look back at how `lonlat.sf` was created. What happened to the `longitude` and `latitude`
   columns, and why does that explain the error?
3. Fix the code so it plots correctly. (Hint: `climatePts` is an `sf` object.)
4. Bonus: even if `longitude` and `latitude` still existed as columns, why would
   `geom_point(aes(x = , y = ))` still be the wrong tool for plotting an `sf` object?

**Report back in the shared doc:** your fixed code, plus a one-sentence explanation of the bug.
