---
title: "Introduction to Spatial Vector Data in R"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(sf)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
gradethis::gradethis_setup()
tutorial_options(exercise.completion = FALSE)
```

## Introduction to Spatial Vector Data in R

The goal of this tutorial is to teach basic methods for working with vector data in R. In addition to demonstrating how to manipulate spatial data objects, the tutorial demonstrates reading and writing spatial vector data. Most spatial vector data processing is handled using the `sf` package and that is the focus of this tutorial.

### Exercise 1 - Create some data

Let's start by creating some point (x, y) data using two individual numerical objects named *longitude* (the x-coordinate) and *latitude* (the y-coordinate). We'll then combine these two objects into a single matrix called *lonlat*.

```{r point-data, exercise=TRUE, exercise.lines = 15, exercise.eval = T}
# set random number seed
set.seed(809)

# longitude object
longitude <- runif(n=10, min=-120, max=-115)

# latitude object
latitude <- runif(n=10, min=35, max=45)

# combine longitude and latitude into a matrix
lonlat <- cbind(longitude, latitude)

# print the result
print(lonlat)
```

Let's answer a few questions about the previous block of code.

```{r random-number-generation, echo=F}
quiz(
  question("What does `set.seed(809)` do?",
    answer("I have no idea."),
    answer("Plants a tree somewhere to fight climate change."),
    answer("Ensures that the sequence of random numbers generated is the same every time the code is run.", correct = TRUE),
    allow_retry = TRUE),
  question("What does `runif()` do?",
    answer("Runs the line of code if a certain condition is met, i.e., runs - if"),
    answer("Generates random numbers from a uniform distribution.", correct = TRUE),
    answer("I have no idea."),
    allow_retry = TRUE))
```

#### **Write R code to perform the following tasks:** 

1. Print the first few rows of `lonlat`.
2. Print what `class` the `lonlat` object is. 
3. Make a scatter plot of the x-y coordinates in `lonlat`. 

```{r exercise1, exercise = TRUE, exercise.lines = 5, exercise.setup = "point-data"}

```

```{r exercise1-solution}
head(lonlat) # or lonlat[1:5,]
class(lonlat)
plot(lonlat)
```
 
```{r exercise1-code-check}
grade_code()
```

### Exercise 2 - Make `lonlat` into a spatial object

Note that the object we just created is of `class::matrix, array` - neither of which is a *spatial data class*. We can convert the `lonlat` object into a spatial data class using the `st_as_sf` function from the `sf` package - **but** we first need to convert `lonlat` into a data frame.

The `sf` package provides classes for geographic vector data. See Chapter 2 of the [Geocomputation with R](https://geocompr.robinlovelace.net/) (https://geocompr.robinlovelace.net/) book for more details. 

```{r sf-library, exercise=TRUE, exercise.lines = 10, exercise.eval = T, echo=T, message=T}
# load sf
library(sf)

# run these outside of this tutorial
# what vignettes are available?
#vignette(package="sf")

# introduction to sf
#vignette("sf1")
```

```{r spatial-object, exercise=TRUE, exercise.lines = 10, exercise.eval = T, exercise.setup = "point-data"}
# convert lonlat to data frame
lonlat.df <- as.data.frame(lonlat)

# convert lonlat.df to spatial object
lonlat.sf <- sf::st_as_sf(lonlat.df, coords = c("longitude", "latitude"))

# print the result
print(lonlat.sf)
```

Some things to note about `lonlat.sf`:  

* It now has *geometry* and a *geometry type*
* It has slot for a *CRS* (coordinate reference system) - in this case, it is NA because we haven't defined it yet.
* It has a *bbox* (bounding box) - the spatial extent of the object.
* It does not have any data columns associated with it (0 fields).

#### **Write R code to perform the following tasks:**  

1. Print the first few rows of `lonlat.sf`.
2. Print what `class` the `lonlat.sf` object is. 
3. Make a plot of `lonlat.sf`. 

```{r exercise2, exercise = TRUE, exercise.lines = 10, exercise.setup = "spatial-object"}

```

```{r exercise2-solution}
head(lonlat.sf) # or lonlat.sf[1:5,]
class(lonlat.sf)
plot(lonlat.sf)
```
 
```{r exercise2-code-check}
grade_code()
```

Note that the plot of `lonlat.sf` does not have axes.

```{r remove-coords-quiz, echo=F}
quiz(
  question("What happened to the `longitude` and `latitude` columns after converting `lonlat.df` to `lonlat.sf` with `st_as_sf()`?",
    answer("They are still present as regular data columns."),
    answer("They were combined into the `geometry` column and removed as separate columns (the default `remove = TRUE`).", correct = TRUE),
    answer("They were renamed to `x` and `y`."),
    allow_retry = TRUE))
```

### Exercise 3 - Define the CRS (Coordinate Reference System)

We will cover the details of **C**oordinate **R**eference **S**ystems (CRS) in the `CRS-and-Spatial-Transformation` tutorial in this package, but for now, let's just define the CRS for the `lonlat.sf` object. Most of the time, spatial data will come with the CRS already defined, but not always. Or, you may created a spatial object from a spreadsheet and will need to define the CRS. Either way, you need to be diligent about understanding and checking the CRS of your data. **NOTE:** *Defining* the CRS is different from *transforming* the data to a different CRS! Defining the CRS is telling `R` what the CRS is, while transforming the data alters calculates the spatial coordinates defining the positions of the object to new coordinates. More on this later too. 

There are two ways you will normally describe a CRS in `R`. An **`EPSG` code** is a short integer that refers to exactly one fully-specified CRS - this is the everyday shorthand and what you should reach for first. **WKT2** (well-known text) is the full, formal description of that same CRS, and it is what `sf` and `terra` actually store and print internally. You may also encounter an older notation, the **`proj4string`** (e.g. `"+proj=longlat +datum=WGS84"`). It still works as *input*, but it has been legacy since PROJ 6 and can silently lose datum information, so prefer EPSG codes or WKT2 for new work. 

We will use the `st_crs` function to define the CRS as EPSG:4326. Look codes up at [epsg.org](https://epsg.org/) (the official IOGP registry) or at [epsg.io](https://epsg.io/4326) (a friendlier third-party front end to the same database). 

```{r crs-1, exercise=TRUE, exercise.eval = T, exercise.lines = 10, exercise.setup = "spatial-object"}
# get attributes of the CRS
sf::st_crs("EPSG:4326")
```

```{r crs-2, exercise=TRUE, exercise.eval = T, exercise.lines = 10, exercise.setup = "spatial-object"}
# assign the CRS for lonlat.sf
sf::st_crs(lonlat.sf) <- 4326

# print the result
print(lonlat.sf)
```

```{r crs-quiz, echo=F}
quiz(
  question("What has changed about the `lonlat.sf` object after defining the CRS?",
    answer("Nothing."),
    answer("The `CRS` is now defined as `EPSG:4326`, which refers to WGS84.", correct = TRUE),
    answer("It now has data columns associated with it."),
    allow_retry = TRUE))
```
    
### Exercise 4 - Adding attribute data to spatial objects

We can add attribute data to spatial objects by creating a data frame and then joining it to a spatial object. The `sf` data class is very flexible, allowing you to interact with attribute data like you would any other data frame.

Let's create a data frame called `climate` with two columns: `temperature` and `precip`. We will then join this data frame to the `lonlat.sf` object.

```{r attribute-data, exercise=TRUE, exercise.eval = T, exercise.lines = 10, exercise.setup = "spatial-object"}
# create the data frame
climate <- data.frame(temperature = rnorm(n=10, mean=25, sd=5), precip = runif(10, min=0, max=500))

# join the climate data to lonlat.sf
climatePts <- lonlat.sf %>%
  dplyr::mutate(climate)

# print the result
print(climatePts)
```

Now, let's plot the `climatePts` object using `ggplot2`. We will use the `temperature` attribute to color the points and the `precip` attribute to size the points.

```{r ggplot-of-climatePts, exercise=TRUE, exercise.eval = T, exercise.lines = 10, exercise.setup = "attribute-data"}
# ggplot2
library(ggplot2)

# plot the result - use geom_sf() since climatePts is an sf object
ggplot(climatePts) +
  geom_sf(aes(color = temperature, size = precip)) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(1, 10)) +
  theme_minimal()
```

### Exercise 5 - Create a polygon geometry

The `sf` data class supports all 17 geometry types defined by the Open Geospatial Consortium, but in practice you will use a handful of them (and none of them are raster classes!): points, lines, polygons, their respective ‘multi’ versions, which group features of the same type into a single feature, and geometry collections, which can hold multiple geometry types in one object.

Here, we create a polygon geometry using the `st_polygon` function. We then convert the polygon geometry to an `sf` object using the `st_sf` function.


```{r polygon-geometry, exercise=TRUE, exercise.lines = 20, exercise.eval = T}
# Define coordinates for the vertices of a polygon
# Note that the polygon must be closed by repeating the first point at the end
x <- c(0, 2, 5, 8, 9, -5, -3, 0)
y <- c(0, 3, 7, 8, 11, 3, 1, 0)

# Combine the x and y coordinates into a matrix
coords <-cbind(x, y)

# Create a polygon geometry
poly_geom <- sf::st_polygon(list(coords))

# Convert to an sf object
poly.sf <- sf::st_sf(geometry = st_sfc(poly_geom))

# Print the result
print(poly.sf)
#
# Plot the result
plot(poly.sf)
```

```{r polygon-closed-quiz, echo=F}
quiz(
  question("Why must the first and last coordinate pairs in `x` and `y` be identical when defining a polygon with `st_polygon()`?",
    answer("They don't need to match; `sf` closes the ring automatically."),
    answer("A polygon geometry must form a closed ring - the boundary has to start and end at the same point.", correct = TRUE),
    answer("It's only required for `multipolygon` geometries, not simple polygons."),
    allow_retry = TRUE))
```

### Exercise 6 - Reading / writing vector shapefiles.

While it is possible to create spatial vector data from scratch, which you may have to do if you are given a spreadsheet with coordinates, it is more common to read in spatial data from an existing shapefile (`*.shp`). Here we will load an example file provided with the `sf` package.  

```{r read-shapefile, exercise=TRUE, exercise.eval = T}
# read the shapefile using st_read
nc <- sf::st_read(dsn = "data/nc.shp")

# print nc
print(nc)
```

Let's plot `nc` and make a map of the `AREA` attribute.

```{r ggplot-of-nc, exercise=TRUE, exercise.eval = T, exercise.lines = 6, exercise.setup = "read-shapefile"}
ggplot(nc) +
  geom_sf(aes(fill = AREA)) +
  theme_minimal()
```

Now let's extract a subset of the data based on the `AREA` attribute and save a new shapefile.

```{r subset-nc, exercise=TRUE, exercise.eval = T, exercise.setup = "read-shapefile"}
# extract the AREA column
nc.area <- nc %>% 
  dplyr::select(AREA)

# print the result
print(nc.area)

# save the subset as a new shapefile
# written to tempdir() so the tutorial does not write into your working directory
sf::st_write(nc.area, file.path(tempdir(), "nc_area.shp"), delete_layer=T)
```

```{r crs-missing-quiz, echo=F}
quiz(
  question("If you read in a shapefile that is missing its `.prj` file, what would `st_crs(nc)` most likely return?",
    answer("An error - `sf` requires a `.prj` file to read a shapefile."),
    answer("`NA` - the CRS is undefined, even though the coordinates themselves are still readable.", correct = TRUE),
    answer("It automatically assumes WGS84 (EPSG:4326)."),
    allow_retry = TRUE))
```

The shapefile is the format you will meet most often, but it is not the only one, and it has real
limitations - notably that it is several files that must travel together, and that it truncates
attribute names to 10 characters. `st_read()` and `st_write()` use GDAL under the hood, so they
handle many other formats, inferred from the file extension. GeoJSON (`.geojson`) is a common
alternative, especially for web mapping, and keeps everything in a single plain-text file.

```{r write-geojson, exercise=TRUE, exercise.eval = T, exercise.lines = 12, exercise.setup = "read-shapefile"}
# same object, written to a different format - just change the extension
gj <- file.path(tempdir(), "nc.geojson")
sf::st_write(nc, gj, delete_dsn = TRUE, quiet = TRUE)

# read it back
nc.gj <- sf::st_read(gj, quiet = TRUE)

# note that the attribute names survive intact (no 10-character truncation)
names(nc.gj)
```

### Exercise 7 - Spatial joins and distances

Two operations come up constantly once you are working with real data: asking *which polygon
does each point fall in?* and *how far apart are these features?*

**Spatial joins.** `st_join()` attaches the attributes of one layer to another based on their
spatial relationship - for example, tagging each occurrence record with the county, watershed, or
protected area it falls inside. It works like a regular table join, except that the matching is
done by geometry rather than by a shared ID column.

```{r spatial-join, exercise=TRUE, exercise.eval = T, exercise.lines = 25, exercise.setup = "read-shapefile"}
# make 20 random points within the bounding box of nc
set.seed(809)
bb <- sf::st_bbox(nc)
pts <- data.frame(lon = runif(20, bb["xmin"], bb["xmax"]),
                  lat = runif(20, bb["ymin"], bb["ymax"]))

# convert to sf, using the SAME CRS as nc (a spatial join requires matching CRSs)
pts.sf <- sf::st_as_sf(pts, coords = c("lon", "lat"), crs = sf::st_crs(nc))

# which county does each point fall in?
pts.joined <- sf::st_join(pts.sf, nc[, c("NAME", "AREA")])

# each point now carries the attributes of the county containing it
head(pts.joined)

# points that fell outside every county get NA
sum(is.na(pts.joined$NAME))

# plot: counties, with the points on top
plot(sf::st_geometry(nc), border = "grey70")
plot(sf::st_geometry(pts.sf), add = TRUE, pch = 19, col = "red")
```

**Distances.** `st_distance()` returns pairwise distances between features. Distance is one of
those calculations where the CRS really matters, so we project first - `nc` is in a geographic
CRS (degrees) and we want an answer in meters.

```{r spatial-distance, exercise=TRUE, exercise.eval = T, exercise.lines = 25, exercise.setup = "read-shapefile"}
# nc is in a geographic CRS
sf::st_is_longlat(nc)

# project to NAD83 / North Carolina (EPSG:32119), a projected CRS in meters
nc.proj <- sf::st_transform(nc, 32119)

# centroids of the first 5 counties
cent <- sf::st_centroid(nc.proj[1:5, ])

# pairwise distances among those 5 centroids - note the units attached to the result
d <- sf::st_distance(cent)
d

# distance from the first county centroid to every county, converted to km
d1 <- sf::st_distance(cent[1, ], nc.proj)
round(sort(as.numeric(d1))[1:5] / 1000, 1)

# which other county is nearest to the first one?
nearest <- sf::st_nearest_feature(cent[1, ], nc.proj[-1, ])
nc.proj[-1, ][nearest, ]$NAME
```

```{r join-distance-quiz, echo=F}
quiz(
  question("You run `st_join(points, polygons)` and every row comes back with `NA` for the polygon attributes, with no error. What is the most likely cause?",
    answer("The polygons layer has no attribute columns to join"),
    answer("The two layers are in different CRSs, or simply don't overlap in space, so nothing matches", correct = TRUE),
    answer("`st_join()` only works on projected data"),
    allow_retry = TRUE),
  question("Why did we project `nc` before calling `st_distance()`?",
    answer("`st_distance()` refuses to run on geographic coordinates"),
    answer("So distances come back in meaningful, constant linear units rather than depending on angular degree coordinates", correct = TRUE),
    answer("Projecting makes the calculation run faster"),
    allow_retry = TRUE))
```
