---
title: "Breakout Activity - Debug: Mismatched CRS"
output: html_document
---

## Breakout Activity - Debug: Mismatched CRS (CRS and Spatial Transformation)

**Groups of 3-4. 5-8 minutes, then brief share-out.**

A colleague is checking whether London falls within the UK boundary polygon. They load the UK
boundary shapefile and reuse the projected London point from earlier in the tutorial, but the
code below throws an error. Work through it as a group, diagnose the error, and fix it.

```{r crs-mismatch-setup}
# recreates the objects from the CRS-and-Spatial-Transformation tutorial, so this chunk runs standalone
library(sf)
library(dplyr)

# UK boundary, in WGS84 (geographic)
uk <- sf::st_read("data/uk.shp", quiet = TRUE)

# London point, transformed to British National Grid (projected) earlier in the tutorial
london <- data.frame(lon = -0.1, lat = 51.5) %>%
  sf::st_as_sf(coords = c("lon", "lat"))
london_geo <- sf::st_set_crs(london, value = 4326)
london_proj <- sf::st_transform(london_geo, crs = 27700)
```

```{r broken-intersects, error=TRUE}
# does the London point fall within the UK boundary?
sf::st_intersects(uk, london_proj)
```

### Discussion questions

1. What error do you get when you run the `broken-intersects` chunk?
2. Check `sf::st_crs(uk)` and `sf::st_crs(london_proj)`. What's different between them, and why
   does that explain the error? (Hint: re-read critical point #2 near the top of the tutorial.)
3. Fix the code so `st_intersects()` runs. There are two valid fixes (transform `uk` to 27700, or
   transform `london_proj` to 4326) — does it matter which one you pick here? Would your answer
   change if `uk` were a raster instead of a polygon? (Hint: re-read critical point #4 near the
   top of the tutorial.)
4. Bonus: `sf::st_set_crs()` would also make the CRS "match" without an error. Why would that be
   the *wrong* fix here?

**Report back in the shared doc:** your fixed code, plus a one-sentence explanation of the bug.
