---
title: "Breakout Activity - Debug: Stacking Mismatched Rasters"
output: html_document
---

## Breakout Activity - Debug: Stacking Mismatched Rasters (Introduction to Raster Data in R)

**Groups of 3-4. 5-8 minutes, then brief share-out.**

A colleague has `myRast` from Exercise 1 of the tutorial, and a second raster covering the same
geographic area but pulled from a different data source. They try to stack the two rasters
together and the code below throws an error. Work through it as a group, diagnose the error, and
fix it.

```{r stack-mismatch-setup}
# recreates the objects from Exercise 1 of the Introduction to Rasters tutorial, so this chunk runs standalone
library(terra)

# myRast, as created in Exercise 1: 10 rows x 10 cols over the same extent used throughout the tutorial
myRast <- terra::rast(ncol = 10, nrow = 10, xmax = -80, xmin = -150, ymin = 20, ymax = 60)
values(myRast) <- 1:ncell(myRast)

# a second raster of the same region, from a different source at a coarser resolution
myRast_other_source <- terra::rast(ncol = 8, nrow = 8, xmax = -80, xmin = -150, ymin = 20, ymax = 60)
values(myRast_other_source) <- 1:ncell(myRast_other_source)
```

```{r broken-stack, error=TRUE}
# stack the two rasters together
rrr <- c(myRast, myRast_other_source)
rrr
```

### Discussion questions

1. What error do you get when you run the `broken-stack` chunk?
2. Print `myRast` and `myRast_other_source` (just type the object name). What's different about
   them, and why does that explain the error? (Hint: re-read the paragraph introducing Exercise 3
   in the tutorial — what does it say a set of rasters needs in common to be stackable?)
3. `myRast` and `myRast_other_source` cover the exact same spatial extent. Does that mean they
   have the same resolution? Why or why not?
4. Fix the code so the two rasters can be stacked. (Hint: look at `terra::resample()` or
   `terra::project()` with a template raster — which one should be resampled to match the other,
   and does it matter which direction you go?)

**Report back in the shared doc:** your fixed code, plus a one-sentence explanation of the bug.
